---
title: 'Codebook'
format:
  html:
    css: styles.css
    embed-resources: TRUE
    code-fold: true
    page-layout: full
    fig_caption: yes
    toc: TRUE
---

This page provides definitions and context for each variable in the CalFresh application dataset used in the Code for America data science exercise.

## About This Dataset

This dataset comes from GetCalFresh.org, a service that helps Californians apply for CalFresh (SNAP) benefits online. The records reflect roughly 2,000 applications submitted in San Diego County, including applicant-submitted data, survey follow-up responses, and county-reported outcomes.

Applicants must:

- Meet income eligibility based on federal poverty thresholds (adjusted for household size).
- Complete a required interview (by phone or in-person).
- Submit verification documents either with their application or afterward.

Note: This data reflects what Code for America collects through GetCalFresh. It does not capture everything a county caseworker uses to make eligibility decisions (e.g., some documents may be mailed or interviews may happen later).

The goal of the associated exercise is to identify factors associated with CalFresh approval, and consider how application design or support could be improved based on the findings.

```{r setup, include=FALSE}
source("R/setup.R")

# Import exercise data
exercise_data <- read_csv("D:/code_for_america/exercise_data.csv")
```

## Codebook

Use the table below to look up definitions and context for each variable in the dataset. This may be useful when checking for missing data or exploring the data structure.

```{r databook}
# Add descriptions to variables
var_desc <- list(
  app_id               = "Unique identifier for each application",
  completion_time_mins = "Time taken to complete the application, in minutes",
  household_size       = "Number of people applying for CalFresh in the household",
  income               = "Total household income in the last 30 days (randomized slightly for privacy)",
  docs_with_app        = "Count of verification documents uploaded with the initial application",
  docs_after_app       = "Count of verification documents uploaded after application (via Later Docs)",
  under18_n            = "Number of children age 17 or younger included in the application",
  over_59_n            = "Number of adults age 60 or older included in the application",
  stable_housing       = "TRUE if applicant rents or owns the place they sleep; FALSE otherwise",
  had_interview        = "TRUE if applicant reported completing the required interview; may be missing",
  zip                  = "ZIP code where the applicant lives or stays",
  approved             = "TRUE if the application was approved for CalFresh by the county"
)

# Use custom R package built by Mari Roberts
databookR::databook(exercise_data, var_descriptions = var_desc)
```

